A Step-Indexed Logic in Agda

### 2023-07-05 papers/cpp2024/paper.tex
progress
check
trying to organize
a little progress

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}
\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
\newcommand{\dotspace}{.\,}
\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\dyn}{\star}
\newcommand{\Dyn}{\ensuremath{\dyn}}
\newcommand{\Int}{\key{int}}
\newcommand{\Bool}{\key{bool}}
\newcommand{\lam}[1]{\lambda #1 \dotspace}
\newcommand{\app}{\,}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\inj}[2]{#1 \,!\, #2}
\newcommand{\proj}[2]{#1 \,?\, #2}

\usepackage{newunicodechar}
\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
\newunicodechar{←}{\ensuremath{\mathnormal\from}}
\newunicodechar{→}{\ensuremath{\mathnormal\to}}
\newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
\newunicodechar{∃}{\ensuremath{\exists}}
\newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
\newunicodechar{⊤}{\ensuremath{\mathnormal\top}}
\newunicodechar{∷}{\ensuremath{\mathnormal::}}
\newunicodechar{ℕ}{\ensuremath{\mathbb N}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{∸}{\ensuremath{\div}}
\newunicodechar{′}{\ensuremath{'}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{₁}{\ensuremath{_1}}
\newunicodechar{₂}{\ensuremath{_2}}
\newunicodechar{Σ}{\ensuremath{\Sigma}}
\newunicodechar{⊎}{\ensuremath{\cupdot}}
\newunicodechar{ᵖ}{\ensuremath{^p}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\not\equiv}}
\newunicodechar{∘}{\ensuremath{\circ}}
\newunicodechar{ᵒ}{\ensuremath{^\circ}}
\newunicodechar{▷}{\ensuremath{\rhd}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{⇔}{\ensuremath{\iff}}
\newunicodechar{⩦}{\ensuremath{\equiv}} % !!
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{∋}{\ensuremath{\ni}}
\newunicodechar{ˢ}{\ensuremath{^s}}
\newunicodechar{ᵈ}{\ensuremath{^d}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{δ}{\ensuremath{\delta}}

@@

%% \author{Lars Th{\o}rv{\"a}ld}
%% \affiliation{%
%%   \institution{The Th{\o}rv{\"a}ld Group}
%%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%%   \city{Hekla}
%%   \country{Iceland}}
%% \email{larst@affiliation.org}

@@

@@

Logical relations are a useful technique for proving properties of
programming languages, from type safety to relational parametricity,
and noninterference.  Initially, logical relations were applied to
terminating languages such as the Simply Typed Lambda
Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}, but
\citet{Appel:2001aa} extended the technique to handle languages with
nontermination by introducing a step index into the logical
relation. After some experience with applications of step-indexed
logical
relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
\citet{Dreyer:2011wl} designed the LSLR logic to streamline the
reasoning about step indexes. This research program eventually
produced Iris~\citep{JUNG:2018aa}, a system that supports a
combination of step indexed logical relations and separation logic
within the Coq proof
assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.

>>>>

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}
\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
\newcommand{\dotspace}{.\,}
\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\dyn}{\star}
\newcommand{\Dyn}{\ensuremath{\dyn}}
\newcommand{\Int}{\key{int}}
\newcommand{\Bool}{\key{bool}}
\newcommand{\lam}[1]{\lambda #1 \dotspace}
\newcommand{\app}{\,}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\inj}[2]{#1 \,!\, #2}
\newcommand{\proj}[2]{#1 \,?\, #2}
\newcommand{\kapprox}{\mathrm{approx}}
\newcommand{\varnow}{\mathsf{var}\mbox{-}\mathsf{now}}
\newcommand{\laters}{\mathsf{laters}}
\newcommand{\Now}{\mathsf{Now}}
\newcommand{\Later}{\mathsf{Later}}
\newcommand{\zero}{\mathsf{zero}}
\newcommand{\suc}[1]{\mathsf{suc}\app#1}
\newcommand{\AgdaMissingDefinition}[1]{#1}

\usepackage{newunicodechar}
\newunicodechar{←}{\ensuremath{\mathnormal\from}}
\newunicodechar{→}{\ensuremath{\mathnormal\to}}
\newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
\newunicodechar{⇓}{\ensuremath{\mathnormal\Downarrow}}
\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
\newunicodechar{∃}{\ensuremath{\exists}}
\newunicodechar{Π}{\ensuremath{\Pi}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{⊣}{\ensuremath{\dashv}}
\newunicodechar{⊧}{\ensuremath{\models}}
\newunicodechar{⊨}{\ensuremath{\models}}
\newunicodechar{↦}{\ensuremath{\mapsto}}
\newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
\newunicodechar{⊤}{\ensuremath{\mathnormal\top}}
\newunicodechar{∷}{\ensuremath{\mathnormal::}}
\newunicodechar{ℕ}{\ensuremath{\mathbb N}}
\newunicodechar{ℤ}{\ensuremath{\mathbb Z}}
\newunicodechar{𝒫}{\ensuremath{\mathcal P}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{×}{\ensuremath{\times}}
\newunicodechar{∸}{\ensuremath{\minus}}
\newunicodechar{′}{\ensuremath{'}}
\newunicodechar{₁}{\ensuremath{_1}}
\newunicodechar{₂}{\ensuremath{_2}}
\newunicodechar{₃}{\ensuremath{_3}}
\newunicodechar{ⱼ}{\ensuremath{_j}}
\newunicodechar{⊎}{\ensuremath{\cupdot}}
\newunicodechar{∪}{\ensuremath{\cup}}
\newunicodechar{ᵖ}{\ensuremath{^p}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\not\equiv}}
\newunicodechar{∘}{\ensuremath{\circ}}
\newunicodechar{ᵒ}{\ensuremath{^\circ}}
\newunicodechar{▷}{\ensuremath{\rhd}}
\newunicodechar{◁}{\ensuremath{\lhd}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{⇔}{\ensuremath{\iff}}
\newunicodechar{⩦}{\ensuremath{\equiv}} % !!
\newunicodechar{∋}{\ensuremath{\ni}}
\newunicodechar{ˢ}{\ensuremath{^s}}
\newunicodechar{ᵈ}{\ensuremath{^d}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{δ}{\ensuremath{\delta}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
\newunicodechar{Λ}{\ensuremath{\mathnormal\Lambda}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{ϕ}{\ensuremath{\phi}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{þ}{\ensuremath{\th}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{Σ}{\ensuremath{\Sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{♯}{\ensuremath{\sharp}}
\newunicodechar{♭}{\ensuremath{\flat}}
\newunicodechar{ℓ}{\ensuremath{\ell}}
\newunicodechar{ₒ}{\ensuremath{_o}}
\newunicodechar{ᵃ}{\ensuremath{^a}}
\newunicodechar{ⁱ}{\ensuremath{^i}}
\newunicodechar{∎}{\ensuremath{\qed}}

@@

\author{Philip Wadler}
\affiliation{%
  \institution{University of Edinburgh}
  %% \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Edinburgh}
  \country{Scotland}}
%% \email{larst@affiliation.org}

\author{Peter Thiemann}
\affiliation{%
  \institution{University of Freiburg}
  \city{Freiburg}
  \country{Germany}
}

@@

\clearpage

\tableofcontents

@@

Logical relations are a useful technique for proving properties of
programming languages, from type safety to relational parametricity
and noninterference.  Initially, logical relations were applied to
terminating languages such as the Simply Typed Lambda
Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}, but
\citet{Appel:2001aa} extended the technique to handle languages with
nontermination by introducing a step index into the logical
relation. After some experience with applications of step-indexed
logical
relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
\citet{Dreyer:2011wl} designed the LSLR logic to streamline the
reasoning about step indexes. Their research program eventually
produced Iris~\citep{JUNG:2018aa}, a system that supports a
combination of step indexed logical relations and separation logic
within the Coq proof
assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.

TODO: discuss guarded cubical agda
